---
title: "Data gathering and (pre-)processing"
author: "Ian Ondo"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: yes
    keep_tex: yes
    includes:
      in_header: preamble.tex
  html_document: null
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{subcaption}
  - \usepackage{caption}
  - \floatplacement{figure}{H}

fig_caption: yes
keep_tex: yes
vignette: >
  %\VignetteIndexEntry{Data gathering and (pre-)processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
    
bibliography: my_bibtex.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos='H'
)
library(kableExtra)
library(scales)
library(tibble)
options(knitr.kable.NA = '**')
options("rgdal_show_exportToProj4_warnings"="none")
```

:::: {.warningbox data-latex=''}
**Package notes:**  
We need to install the following set of R packages to successfully run the codes in this vignette:

- **UsefulPlants** [@data_table]
- **rWCVP** [@rwcvp]
::::

# Introduction
In this vignette, we will show how to compile and format occurrence data from various data sources available online, or as zipped (compressed) files using **UsefulPlants** and its companion package **rsdm**. We will focus on two online databases:

- *Global Biodiversity Information Facility* (**GBIF**, [*\underline{https://www.gbif.org}*](https://www.gbif.org))

- *Botanical information and Ecology Network* (**BIEN version 4.1**, [*\underline{https://bien.nceas.ucsb.edu/bien}*](https://bien.nceas.ucsb.edu/bien)) 

Both have R packages facilitating the download of occurrence records directly from R. For example, the function *get_gbifid_* from the R package **taxize** [@taxize] and function *occ_download* from the R package **rgbif** [@rgbif] can be used for matching species names and downloading occurrence records from GBIF respectively, while the function *BIEN_occurrence_species* from the R package **BIEN** [@BIEN]helps to extract occurrence records from the BIEN database.  
To our knowledge, other databases such as **RAINBIO** ([\underline{http://rainbio.cesab.org/}](http://rainbio.cesab.org/)), **speciesLink** ([\underline{http://www.splink.org.br/}](http://www.splink.org.br/)), **BioTIME** ([\underline{https://biotime.st-andrews.ac.uk/}](https://biotime.st-andrews.ac.uk/)) or **Genesys** ([\underline{https://www.genesys-pgr.org}](https://www.genesys-pgr.org)) do not provide R packages to access their data, but snapshots of the entire database can be accessed, sometimes on request, and often shared as a (compressed) file. In this case the **data.table** [@data_table] is a very useful R package in terms of efficiency and speed to process these type of datasets.

# I. Download a list of useful plants species names
Download the useful plants database used in @UsefulPlants [\textcolor{blue}{\underline{here}}](https://www.here.com), and unzip it into a folder, e.g. *"usefulplant_data"*. Then, read in the data and filter as needed:
```{r load-useful-plant-data, eval=FALSE, include=TRUE}

# load the data into memory
useful_plant_db <- data.table::fread(input="usefulplant_data/useful_plants_data_list.csv",
                                     h=TRUE,
                                     showProgress=FALSE)

useful_plant_db

# get the list of all useful plants names in the database
useful_plant_names <- useful_plant_db %>%
  dplyr::pull(binomial_acc_name) # extract binomial accepted names

# optionally filter e.g. edible plants
edible_plant_names <- useful_plant_db %>%
  dplyr::filter(HumanFood==1) %>%         # select edible plants
  dplyr::pull(binomial_acc_name)          
```

 
# II. Download occurrence data from GBIF

**rsdm** has a function called `queryGBIF` that uses both *get_gbifid_* and *occ_download* to obtain occurrence records from GBIF from a list of species names. The function `get_gbiftaxonkey` is used internally to get taxon *keys*, and the essential of the code is described in this blog [*\textcolor{blue}{\underline{downloading a long species list on GBIF}}*](https://data-blog.gbif.org/post/downloading-long-species-lists-on-gbif/), feel free to have a look at it.  
Basically, we first need to create a GBIF account because we are going to use GBIF credentials as explained in the blog to login to GBIF and download occurrence records for a fairly large number of species.

## II.1 *create a GBIF account*
Go to the GBIF website homepage [\underline{https://www.gbif.org}](https://www.gbif.org) \rightarrow click on the `login` button on the top-right corner of the page \rightarrow click on `register` \rightarrow fill up the fields to create your account. Then, save your *\underline{GBIF username}*, *\underline{email}* associated with username and *\underline{password}* for logging in to GBIF somewhere safe, for example in your .Renviron file:
```R
install.packages("usethis") # if not yet installed
library(usethis)

# open your .Renviron file
edit_r_environ()

# write down your GBIF credentials as environmental variables
GBIF_USER = "your_username"
GBIF_EMAIL = "your_email_address"
GBIF_PWD = "your_password"

# save your changes, close the file and restart your R session
```
Check that your GBIF credentials are now accessible by typing in the console e.g. `Sys.getenv("GBIF_USER")`.

## II.2 *use `queryGBIF` to download occurrence records*
```{r queryGBIF, eval=FALSE, message=FALSE, include=TRUE}

# load the rsdm package
library(rsdm)

# make sure species names are correctly formatted 
edible_plant_names <- gsub("_"," ",edible_plant_names)

# specify a folder where to store the zip file of your downloads
# it is recommended to use a specific folder for better traceability
dwnld_dir = tempdir()
  
# run the query using the vector of plant species names
gbifDATA = queryGBIF(species_name=edible_plant_names,
                gbif_download_dir=dwnld_dir,
                # time in seconds before dropping out the connection
                time_out = 300)

```

:::: {.infobox data-latex=''}
By default, **queryGBIF** will return after `time_out` seconds. If GBIF has not finished processing your query or if something went wrong, `queryGBIF` will returns `NULL` and an error message. If this is the case: 

 - **\underline{Login}** to your account on GBIF.
 
 - Go to your downloads and **\underline{check}** out the status of your query.
 
 - Once available, you can manually **\underline{download}** your dataset as zip file in the directory of your choice.
::::

If everything worked fine, the output of `queryGBIF` is a `data.table` object with your occurrence records with the following columns:

- *"species"*

- *"fullname"*,

- *"decimalLongitude"*

- *"decimalLatitude"*

- *"countryCode"*

- *"coordinateUncertaintyInMeters"*

- *"year"*

- *"individualCount"*

- *"gbifID"* 

- *"basisOfRecord"* 

- *"institutionCode"* 

- *"establishmentMeans"*

- *"is_cultivated_observation"*

- *"sourceID"* (always GBIF)

Find documentation on most of the fields [*https://www.gbif.org/data-quality-requirements-occurrences*](https://www.gbif.org/data-quality-requirements-occurrences).

If the automatic download failed and you ended up downloading the zip file manually, you can still format your dataset as follows:
```{r formatting-GBIF-dataset, eval=FALSE, message=FALSE, include=TRUE}

# get the path to your zip file
path_to_my_zip.file <- "path/to/yourzip.file"

# read in the data from your zip file
gbifDATA <- data.table::fread(cmd=paste("unzip -p",path_to_my_zip.file),
                                       header=TRUE,
                                       showProgress=FALSE,
                                       na.strings=c("",NA),
                                       fill=FALSE,
                                       quote = "",
                                       select=c("species",
                                               "taxonRank",
                                               "infraspecificEpithet",
                                               "decimalLongitude",
                                               "decimalLatitude",
                                               "countryCode",
                                               "coordinateUncertaintyInMeters",
                                               "year",
                                               "gbifID",
                                               "basisOfRecord",
                                               "institutionCode",
                                               "establishmentMeans",
                                               "individualCount"))

gbifDATA<-na.omit(gbifDATA, cols= c("decimalLatitude", "decimalLongitude"))

# format the data:
gbifDATA[, `:=`(taxonRank = ifelse(taxonRank %in% c("SPECIES","GENUS"),
                                   NA,
                                   ifelse(taxonRank=="FORM", "f.",
                                          ifelse(taxonRank=="SUBSPECIES", "subsp.",
                                                 ifelse(taxonRank=="VARIETY","var.",
                                                        taxonRank)))),
                    countryCode = countrycode::countrycode(countryCode,
                                                           origin =  'iso2c',
                                                           destination = 'iso3c',
                                                           nomatch = NA),
                    establishmentMeans = ifelse(establishmentMeans=="INTRODUCED",
                                                "Introduced",
                                                ifelse(establishmentMeans== "NATIVE",
                                                       "Native",
                                                       establishmentMeans)))]

# create 'fullname', 'is_cultivated_observation' and 'sourceID' columns
gbifDATA[, `:=`(fullname = ifelse(is.na(infraspecificEpithet),
                                  paste(species),
                                  paste(species, taxonRank, infraspecificEpithet)),
                is_cultivated_observation = NA,
                sourceID = 'GBIF')]

# delete taxonRank and infraspecificEpithet columns
gbifDATA[, `:=`(taxonRank = NULL, infraspecificEpithet = NULL)]

# remove fossil records
gbifDATA <- gbifDATA[basisOfRecord!="FOSSIL_SPECIMEN"]

# set column order
colNames = c("species",
             "fullname",
             "decimalLongitude",
             "decimalLatitude",
             "countryCode",
             "coordinateUncertaintyInMeters",
             "year",
             "individualCount",
             "gbifID",
             "basisOfRecord",
             "institutionCode",
             "establishmentMeans",
             "is_cultivated_observation",
             "sourceID")

data.table::setcolorder(gbifDATA, colNames)

# set the key to the species column to enable fast binary search
data.table::setkey(gbifDATA, 'species')
```
# III. Download occurrence data from BIEN

Similarly, **rsdm** has a function called `queryBIEN` to download occurrence records from the BIEN database. Internally, `queryBIEN` uses `BIEN_occurrence_species` with the following default arguments:

```R
cultivated = TRUE,
only.new.world = FALSE,
all.taxonomy = TRUE,
native.status = TRUE,
observation.type = TRUE,
political.boundaries = TRUE,
natives.only = FALSE,
collection.info = TRUE
```

Check out the meaning of these arguments from the package help here 
[*\textcolor{blue}{\underline{manual}}*](https://search.r-project.org/CRAN/refmans/BIEN/html/BIEN_occurrence_species.html)

```{r queryBIEN, eval=FALSE, message=FALSE, include=TRUE}

# run the query using the vector of plant species names previously defined in section II.2 
bienDATA = queryBIEN(species_name=edible_plant_names)

```

If everything workec fine, `bienDATA` should be a `data.table` object formatted as `gbifDATA`, i.e. with the same column names.

# IV. Formatting occurrence data from *RAINBIO*, *speciesLink*, *BioTIME* and *Genesys*
For consistency between the different databases, we need to format each occurrence dataset the same way as we did for *GBIF* and *BIEN*.  
Let's assume that each dataset comes as a zipped folder, if this is not the case, replace `paste("unzip - p","path/to/file.zip")` by the path to your unzipped file in the relevant part of the code.

```R
# define column names
colNames = c("species",
             "fullname",
             "decimalLongitude",
             "decimalLatitude",
             "countryCode",
             "coordinateUncertaintyInMeters",
             "year",
             "individualCount",
             "gbifID",
             "basisOfRecord",
             "institutionCode",
             "establishmentMeans",
             "is_cultivated_observation",
             "sourceID")
```
## *RAINBIO*
```{r formatting-rainbio, eval=FALSE, include=TRUE}

# read in the data
rainbioDATA <- data.table::fread(paste("unzip -p","path/to/file.zip"),
                                     header=TRUE,
                                     showProgress=FALSE,
                                     select=c("tax_sp_level",
                                      "species",
                                      "decimalLatitude",
                                      "decimalLongitude",
                                      "iso3lonlat",
                                      "basisOfRecord",
                                      "institutionCode",
                                      "catalogNumber",
                                      "coly"))

#------------
#= formatting
#------------
# change the column names
data.table::setnames(rainbioDATA, 
                     c("species",
                       "fullname",
                       "decimalLatitude",
                       "decimalLongitude",
                       "countryCode",
                       "basisOfRecord",
                       "institutionCode",
                       "gbifID",
                       "year"))

# remove NA coordinates
rainbioDATA <- na.omit(rainbioDATA, cols= c("decimalLatitude", "decimalLongitude"))

# create additional columns: 'coordinateUncertaintyInMeters' , 'is_cultivated_observation',
# 'establishmentMeans', 'individualCount' and 'sourceID'

rainbioDATA[ ,`:=`(coordinateUncertaintyInMeters = NA,
                  is_cultivated_observation = "No",
                  establishmentMeans = NA,
                  individualCount = NA,
                  sourceID = 'RAINBIO')]

# set column order
data.table::setcolorder(rainbioDATA, colNames)

# set the key to the species column to enable fast binary search
data.table::setkey(rainbioDATA, 'species')

```

## *speciesLink*
```{r formatting-speciesLink, eval=FALSE, include=TRUE}

# read in the data
spLinkDATA <- data.table::fread(paste("unzip -p","path/to/file.zip"),
                                    header=TRUE,
                                    showProgress=FALSE,
                                    encoding = 'UTF-8',
                                    na.strings=c("",NA),
                                    select= c("query",
                                             "scientificname",
                                             "longitude",
                                             "latitude",
                                             "country",
                                             "coordinateprecision",
                                             "yearcollected",
                                             "individualcount",
                                             "catalognumber",
                                             "basisofrecord",
                                             "institutioncode"))

#------------
#= formatting
#------------
# change the column names
data.table::setnames(spLinkDATA, 
                     c("species",
                       "fullname",
                       "decimalLatitude",
                       "decimalLongitude",
                       "countryCode",
                       "coordinateUncertaintyInMeters",
                       "year",
                       "individualCount",
                       "gbifID",
                       "basisOfRecord",
                       "institutionCode"))

# remove NA coordinates
spLinkDATA <- na.omit(spLinkDATA, cols= c("decimalLatitude", "decimalLongitude"))

spLinkDATA[ ,`:=`(species = paste0(substr(species,1,1),
                                   tolower(substr(species,2,nchar(species)))),
                 countryCode = ifelse(countryCode=='Brasil','Brazil',countryCode),
                 basisOfRecord = ifelse(basisOfRecord=="S","SPECIMEN",
                                        ifelse(basisOfRecord== "O",
                                               "OBSERVATION",
                                               basisOfRecord)),
                 is_cultivated_observation = NA,
                 establishmentMeans = NA,
                 sourceID = 'spLink')]

spLinkDATA[ ,countryCode:= countrycode::countrycode(countryCode,
                                                    origin =  'country.name',
                                                    destination = 'iso3c',
                                                    nomatch = NA)]
# set column order
data.table::setcolorder(spLinkDATA, colNames)

# set the key to the species column to enable fast binary search
data.table::setkey(spLinkDATA, 'species')

```

## *bioTIME*
```{r formatting-biotime, eval=FALSE, include=TRUE}

# read in the data
biotimeDATA  <- data.table::fread(paste("unzip -p","path/to/file.zip"),
                                  header=TRUE,
                                  showProgress=FALSE,
                                  select=c("GENUS_SPECIES",
                                           "LATITUDE",
                                           "LONGITUDE",
                                           "YEAR",
                                           "sum.allrawdata.ABUNDANCE",
                                           "SAMPLE_DESC"))

#---------------
#= formatting
#---------------

# change the column names
data.table::setnames(biotimeDATA, c("species",
                                    "decimalLatitude",
                                    "decimalLongitude",
                                    "year",
                                    "individualCount",
                                    "gbifID"))

# remove NA coordinates
biotimeDATA  <- na.omit(biotimeDATA, cols= c("decimalLatitude", "decimalLongitude"))

# create additional columns: 'fullname', 'coordinateUncertaintyInMeters','institutionCode',
#'countryCode', 'basisOfRecord', 'establishmentMeans', 'is_cultivated_observation'
# and 'sourceID'
biotimeDATA[ ,`:=`(fullname = species,
                   coordinateUncertaintyInMeters = NA,
                   institutionCode = NA,
                   countryCode = NA,
                   basisOfRecord = NA,
                   establishmentMeans = NA,
                   is_cultivated_observation = "No",
                   sourceID = 'BIOTIME')]

# set column order
data.table::setcolorder(biotimeDATA, colNames)

# set the key to the species column to enable fast binary search
data.table::setkey(biotimeDATA, 'species')

```

## *Genesys*
```{r formatting-genesys, eval=FALSE, include=TRUE}

# read in the data
genesysDATA <- data.table::fread(paste("unzip -p","path/to/file.zip"),
                                     header=TRUE,
                                     showProgress=FALSE,
                                     na.strings=c("",NA),
                                     select= c("GENUS",
                                    "SPECIES",
                                    "SUBTAXA",
                                    "DECLONGITUDE",
                                    "DECLATITUDE",
                                    "ORIGCTY",
                                    "COLLSRC",
                                    "INSTCODE",
                                    "UUID",
                                    "COLLDATE",
                                    "COORDUNCERT",
                                    "SAMPSTAT"))

#---------------
#= formatting
#---------------

# change the column names
data.table::setnames(genesysDATA,
                     c("genus",
                       "species",
                       "subtaxa",
                       "decimalLongitude",
                       "decimalLatitude",
                       "countryCode",
                       "basisOfRecord",
                       "institutionCode",
                       "gbifID",
                       "year",
                       "coordinateUncertaintyInMeters",
                       "is_cultivated_observation"))

# format the data: 
# concatenate genus+species,
# create full species name,
# extract the year of the collection,
# transform information about cultivated observation and basis of records in categorical variable
genesysDATA[, `:=`(species = paste(genus, species),
                  fullname = ifelse(is.na(subtaxa),
                                    paste(genus, species),
                                    paste(genus, species, subtaxa)),
                  year = as.numeric(stringr::str_extract(year,"[[:digit:]]{4}")),
                  is_cultivated_observation = ifelse(is_cultivated_observation==999,
                                                     NA,
                                                     ifelse(is_cultivated_observation > 300,
                                                            "Yes",
                                                            "No")),
                  basisOfRecord = cut(basisOfRecord, 
                                      breaks=c(10, 16, 29, 59, 63),
                                      labels=c(
                                        "Wild habitat",
                                        NA,
                                        "Cultivated habitat",
                                        "Wild habitat")
                                      ))]

# create additional columns: 'establishmentMeans', 'individualCount' and 
# 'sourceID' and delete 'genus' and 'subtaxa'
genesysDATA[ ,`:=`(genus = NULL,
                  subtaxa = NULL,
                  establishmentMeans = NA,
                  individualCount = NA,
                  sourceID = 'GENESYS')]

# set column order
data.table::setcolorder(genesysDATA, colNames)

# set the key to the species column to enable fast binary search
data.table::setkey(genesysDATA, 'species')

```

:::: {.warningbox data-latex=''}
**Resolving species taxonomy**
In vignette, we showed how to obtain occurrence records from a list of species names. issued from the World Checklist of Vascular Plants (WCVP) [@WCVP]. Since, plant taxonomy backbone often differ between databases, we need to make sure that 
::::

# References
